### 2022-06-10 paper.tex
minor intro adjustments

To detect CpG methylation by sequencing, bisulfite conversion is commonly employed: treatment of DNA with bisulfite causes unmethylated cytosine bases to be deaminated to uracil, which is paired with thymine in subsequent PCR. Therefore, unmethylated CpGs are read as TG in sequencing of bisulfite-converted DNA while only methylated CpG is correctly read as CG \citep{Frommer_1992}. Such so-called bisulfite sequencing has been used for bulk samples since long, and methods for bisulfite sequencing with single-cell resolution (scBS) were recently developed \citep{Smallwood_2014} and are increasingly used.

CpG dinocleotides are underrepresented in eukaryotic genomes (due to their vulnerability to mutation by conversion to TpG due to accidental deamination), and the majority of existing CpG sites are methylated. Presence or absence of the methyl group at CpG sites is believed to have effect on gene regulation, with methylation inhibiting transcription. Methylating DNA during differentiation is therefore conjectured to be an important mechanism to silence genome regions that are not needed in the target lineage. One important application of single-cell bisulfite sequencing (scBS) therefore lies in the study of differentiation processes.

@@

This procedure is commonly adapted when working with single-cell DNA-methylation data, so that all the tools used in scRNA-Seq downstream of the PCA step can be used. The question is therefore how to construct a suitable matrix as input for the PCA. A simple and popular approach is to tile the genome into windows of, say, 100 kB size, and calculate for each cell the average methylation of the DNA within each window. To this end, we identify all CpG sites in the window for which we have coverage with at least one read and can hence call the CpG to be either unmethylated or methylated in the cell. Then we denote as average DNA methylation of this window in a given cell the proportion of observed CpG sites in the window have been found to be methylated. This yields a matrix, with one row for each cell and one column for each window from the genome tiling, comprising numbers (methylation fractions) between 0 and 1. This matrix is now subjected to PCA.

As an example, we will us in this paper data from XX et al., who have performed scBS on [...] and analysed their data in the manner just described. Figure \ref{umaps}A shows a UMAP plot, derived from our re-analysis of the data, using a tiling of the genome into windows of size 100 kb. As one can see, the approach is suitable to see clear structure in the data, observe ...

>>>>

To detect CpG methylation by sequencing, bisulfite conversion is commonly employed: treatment of DNA with bisulfite causes unmethylated cytosine bases to be deaminated to uracil, which is replaced by thymine in subsequent PCR. Therefore, unmethylated CpGs are read as TG in sequencing of bisulfite-converted DNA while only methylated CpG sites are correctly read as CG \citep{Frommer_1992}. Such so-called bisulfite sequencing has been used for bulk samples since long, and methods for bisulfite sequencing with single-cell resolution were recently developed \citep{Smallwood_2014} and are increasingly used.

CpG dinocleotides are underrepresented in eukaryotic genomes (due to their vulnerability to mutation by conversion to TpG due to spontaneous deamination), and the majority of existing CpG sites are methylated. Presence or absence of the methyl group at CpG sites is believed to have effect on gene regulation, with methylation inhibiting transcription. The removal of DNA methylation during cell differentiation is therefore conjectured to be an important mechanism to activate genome regions that are required in the target lineage and define cell identity \citet{cedar2022role}. One important application of single-cell bisulfite sequencing (scBS) therefore lies in the study of differentiation processes, for instance in embryonic development \citet{argelaguet2019gastru}.

@@

This procedure is commonly adapted when working with single-cell DNA-methylation data, so that established methods used in scRNA-Seq can be used.
The question is therefore how to construct a suitable matrix as input for the PCA. A simple and common approach, used for instance by \citet{luo2017single}, is to tile the genome into windows of, say, 100~kb size, and calculate for each cell the average methylation of the DNA within each window. To this end, we identify all CpG sites in the window for which we have coverage with at least one read and can hence call the CpG to be either unmethylated or methylated in the cell. Then we denote as average DNA methylation of this window in a given cell the proportion of observed CpG sites in the window that were found to be methylated (Figure \ref{fig:smoothres}A). This yields a matrix, with one row for each cell and one column for each window from the genome tiling, comprising numbers (methylation fractions) between 0 and 1. This matrix is now subjected to PCA.

As an example, we will use in this paper data from \citet{luo2017single}, who performed scBS on 16 neuronal subtypes of the mouse frontal cortex, and analysed their data in the manner just described. Figure \ref{umaps}A shows a UMAP plot, derived from our re-analysis of the data, using a tiling of the genome into windows of size 100~kb. As one can see, the approach is suitable to see clear structure in the data, observe ...
